Explore the factors that may be associated with obesity

### 2024-05-03 products/manuscript/Manuscript copy.qmd
improve after review

[Obesity](https://www.who.int/news-room/fact-sheets/detail/obesity-and-overweight)![]() is a chronic complex disease defined by excessive fat deposits that can impair health.
Overweight and obesity are predominately a result of a sustained positive energy balance, stemming from a combination of excess dietary energy intake (mainly due to poor eating habits) and reduced energy expenditure (due to lack of physical activity and prolonged sedentary activities).
It is a significant risk factor for and contributor to increased morbidity and mortality, mainly by its increasing risk of type 2 diabetes and heart disease, as well as certain types of cancer and chronic diseases, including osteoarthritis, liver and kidney disease, sleep apnea, and depression [@pi-sunyer2009].
Also it does affct bone health and reproduction.

In the recent decades, obesity has become a more and more common and severe problem across the world with worldwide obesity rates tripling since 1975, thus reaching pandemic proportions [@nu15132896].
Worldwide adult obesity has more than doubled since 1990, and adolescent obesity has quadrupled.
In 2022, 1 in 8 people in the world were living with obesity.
In the US.
the situation seems similar with the worldwide trend.
In 2022, 22 states had at least 35% of adults with obesity, up from 19 states in 2021.
Ten years ago, CDC said, no state had an adult obesity prevalence at or above 35%.

Obesity level is basically defined by Body Mass Index ([BMI](https://www.cdc.gov/obesity/basics/adult-defining.html)).
BMI is a person’s weight in kilograms divided by the square of height in meters.
A high BMI can indicate high body fatness.
BMI less than 18.5 is defined as underweight; BMI range between 18.5-24.9 falls within the healthy weight range.
If BMI is from 25 to 29.9, then the person is regarded as overweight.
And when someone's BMI is higher than 30, he/she is at the obesity range.

There is no doubt that obesity is a multifarious disease due to cryogenics environments, psycho-social factors and genetic variants in most cases, but many factors can contribute to excess weight gain including eating patterns, physical activity levels, and sleep routines.
Social determinants of health, genetics, and taking certain medications also play a role.

There is a lot of discussion on the Internet and social media about fitness and weight loss, and people assume that physical activity can lose weight as a fact.
Also KOLs usually encourage people to drink more water for higher metabolic levels.
But I have not seen a clear data to prove it.
At the same time, the effect of physical activity to lose weight is also different from person to person.

@@

The data is obtained from a website [UCI Machine Learning Repository](https://archive.ics.uci.edu/dataset/544/estimation+of+obesity+levels+based+on+eating+habits+and+physical+condition) which have collected the survey results of individuals from the countries of Mexico, Peru and Colombia on their eating habits and physical condition, as well as their estimation of obesity levels.
The original dataset contains 17 variables and 2111 observations, labeled with the class variable Obesity Level which has 7 categories.

@@

I would like to figure out if higher frequency of physical activity and drinking more water could keep people away from obesity and whether heredity is an important factor for obesity.
Also, I will focus on the relationship between alcohol consumption frequency and obesity level.

@@

Then simple linear fit will be applied to explore the relationship between variables and outcomes (both categorical and numeric), and proper predictors will be pick out from the variables.
And then machine learning models will be chosen and fitted to the dataset wieh each performance metrics will be determined and compared for reporting the "best" model among the choices based on reality.

@@

The dataset presents the results of a survey on the topic of obesity including eating habits and personal physical conditions and answers to behavioral questions.

@@

@@

First, I was curious that whether there is a negative correlation between transportation type and physical activity frequency, I converted transportation types into numbers indicating the energy consuming levels from 1 (lowest automobile) to 4 (highest walking).
Then I used correlation function to calculate the coefficient for these two variable but it was pretty low (0.0059).

Then, I would like to fit linear model for BMI by both water consumption and physical activity.
And the result indicates that there is a connection effect between them two, which means that both of them contribution to obesity control.

Also, I use single indicator, alcohol consumption and BMI as an outcome to fit linear model.
It shows that if an individual consumes alcohol sometimes than never, his BMI increase 3.98.
But the correlations are all negative for level 3 and 4, which means consuming more alcohol may decrease BMI.

Because I still wonder if obesity could be related to gender or age.
So I fit simple linear model to the dataset using gender and age as independent variable.
The result indicates the interactions between them.
It is a factor out of my expectation.
And I decide to include this interaction for next machine learning models.

In the last model I select all the variables that I think may have effect on BMI as predictors and BMI as outcome.

Since I think it is quite obvious that family history is an important factor for obesity, and also I have mentioned some reasons how family obesity could influence the kids involved, I choose to skip it for this basic exploratory analysis here.
But I will include it in the following part.

@@

+-----------------------------------+-------------------------------------------+---------------+------------------------------+
| Model                             | Predictor                                 | Metric Value  | Metric Value on testing data |
+===================================+===========================================+===============+==============================+
| Linear null                       | \-                                        | RMSE=8.00     | RMSE=8.04                    |
+-----------------------------------+-------------------------------------------+---------------+------------------------------+
| Linear regression 1               | Age, History, Alcohol, MTRANS             | RMSE=0.62     | RMSE=6.61                    |
+-----------------------------------+-------------------------------------------+---------------+------------------------------+
| Linear regression 2               | All                                       | RMSE=6.30     | RMSE=3.99                    |
+-----------------------------------+-------------------------------------------+---------------+------------------------------+
| Linear regression 2 + LASSO model | All                                       | RMSE=7.38     | RMSE=7.38                    |
|                                   |                                           |               |                              |
|                                   |                                           | l=0.0011      |                              |
+-----------------------------------+-------------------------------------------+---------------+------------------------------+
| Random forest                     | Gender, Age, History, Water, Alcohol, FAF | RMSE=3.22     | RMSE=3.97                    |
+-----------------------------------+-------------------------------------------+---------------+------------------------------+
| Logistic null                     | \-                                        | Accuracy=0.17 | Accuracy=0.15                |
+-----------------------------------+-------------------------------------------+---------------+------------------------------+
| Discriminant analysis             | All                                       | Accuracy=0.50 | Accuracy=0.48                |
+-----------------------------------+-------------------------------------------+---------------+------------------------------+
| Multinominal logistic regression  | All                                       | Accuracy=0.58 | Accuracy=0.52                |
+-----------------------------------+-------------------------------------------+---------------+------------------------------+

The above table has listed all of the the models I have chosen to fit the data together with the null models.

When I use BMI as a continuous outcome, linear regression models can be applied to the data.
Compared to the linear null model, I think all of the ML models perform better according to RMSE value.
Since I think it is reasonable that physical activity is a potential factor that may influence the obesity level, although linear model with all variables and interactions has higher RMSE, I think regression model 2 that picks all of the predictors is better.
Then since random forest model selects all predictors and RMSE value is relatively lower among all the linear models, RF model is the choice for linear model.

@@

![](images/Observed-Predicted .png){width="2100"}

As for obesity level as a categorical outcome, logistic models are fit to the data.
And I use accuracy as the performance metric for this evaluation.
First, both the models I use have better performance than null model.
And according to the accuracy, multinominal logistic regression performs better.

Because I have chosen random forest model between linear regression models and multinominal logistic regression between logistic models, I use test data to assess the quality of these two models separately.
And the new RMSE on test data for random forest model is 3.97, and it is 3.22 when the model is trained and cross validation is applied.
And the new Accuracy is 0.52 when it is used on test data, also lower than 0.58 when the model is trained.
The metric value difference is a sign that both of the models are overfitting.

So I try to use test data to assess all of the models in the table above and collect new metric values.
For linear regression 1, the RMSE on testing data is 6.61, which is much higher than that on training data.
So this model is not doing prediction well.
The new RMSE for linear regression 2 is 3.99 lower than 6.30 on training data, which is a sign of good prediction.
Besides, LASSO model has the same RMSE on both training data and testing data which is 7.38.
It is also a reasonable result for good prediction.
Both logistic models have lower accuracy than on test data than training data.
Based on all the result of the machine learning I have used on dataset, I think linear regression model with LASSO regularization is a good choice among them.

@@

I chose a dataset that summarize an investigation result done in South American countries about individual physical behaviors and the classification of obesity level based on the information of height and weight obtained from the investigation.
I had some questions to explore and did data analysis on this dataset.

@@

For machine learning model fitting, I found the best predictors for BMI index based on the value of some performance metrics after trying both linear regression and classification model, which were gender, age, family history, water drinking amount, frequency of physical activity, alcohol consumption frequency and commute transporation.
Although the machine learning models that I used for fitting the dataset are different types and their metrics and not be compared directly, the assessment of model quality using testing data could give a clue for the model selection.
Considering all the results of performance metrics, I finally chose liner regression model with LASSO regularization with all predictors plus BMI as the outcome.
The reason was that RMSE calculated of predicted MBI values on testing data split from cleaned dataset was as same as the result when it was trained.
Also, regularization was applied by LASSO model to reduce overfitting at the same time.

@@

I think my analysis has include two different types of model according to the type of outcome and each type include 2 models, which can provide more information and the result can be shown in different aspects as well as more choices were provided.
Also, interactions between predictors were found and included in the model fitting process.
Both predictors with and without interactions were all fit and compared with each other.
And LASSO model was used for decreasing overfitting.

I did tried to do tuning for random forest model but I failed for running it successfully, so this could be a limitation that I have.
And maybe there could be more metrics that I can also calculate but I only chose 1 for each model.
Furthermore, chatGPT told me that even if testing result has a sign for well predicting, there still could be other factors for further assessment such as F1-score and so on.
But base on my understanding, I don't know what other methods I could choose from so I just stopped here.

>>>>

[Obesity](https://www.who.int/news-room/fact-sheets/detail/obesity-and-overweight)![]() is a chronic complex disease defined by excessive fat deposits that can impair health.
Overweight and obesity are predominately a result of a sustained positive energy balance, stemming from a combination of excess dietary energy intake (mainly due to poor eating habits) and reduced energy expenditure (due to lack of physical activity and prolonged sedentary activities).
It is a significant risk factor for and contributor to increased morbidity and mortality, mainly by its increasing risk of type 2 diabetes and heart disease, as well as certain types of cancer and chronic diseases, including osteoarthritis, liver and kidney disease, sleep apnea, and depression [@pi-sunyer2009].
Also it does affect bone health and reproduction.

In the recent decades, obesity has become a more and more common and severe problem across the world with worldwide obesity rates tripling since 1975, thus reaching pandemic proportions [@nu15132896].
Worldwide adult obesity has more than doubled since 1990, and adolescent obesity has quadrupled.
In 2022, 1 in 8 people in the world were living with obesity.
In the US.
the situation seems similar with the worldwide trend.
In 2022, 22 states had at least 35% of adults with obesity, up from 19 states in 2021.
Ten years ago, according to [CNN Health](https://www.cnn.com/2023/09/21/health/obesity-more-common-states-cdc-data/index.html), CDC said there was no state had an adult obesity prevalence at or above 35%.

Obesity level is basically defined by Body Mass Index ([BMI](https://www.cdc.gov/obesity/basics/adult-defining.html)).
BMI is a person's weight in kilograms divided by the square of height in meters.
A high BMI can indicate high body fatness.
BMI less than 18.5 is defined as underweight; BMI range between 18.5-24.9 falls within the healthy weight range.
If BMI is from 25 to 29.9, then the person is regarded as overweight.
And when someone's BMI is higher than 30, he/she is at the obesity range.

There is no doubt that obesity is a multifarious disease due to cryogenics environments, psycho-social factors and genetic variants in most cases, but many factors can contribute to excess weight gain including eating patterns, physical activity levels, and sleep routines.
Social determinants of health, genetics, and taking certain medications also play a role.
There is an [article](https://www.nichd.nih.gov/health/topics/obesity/conditioninfo/cause) by NIH listing some causative factors for obesity, including food and activity, environment effect, genetics, health conditions and medications, stress, emotional factors, and poor sleep.
Other [articles](https://ro.co/weight-loss/obesity-causes-risk-factors-strategies/#risk-factors-for-obesity) also have mentioned hormone as cause and family history, age, sedentary lifestyle and smoking as risk factors for obesity.
But alcohol consumption is not that frequently mentioned as far as I know.

There is a lot of discussion on the Internet and social media about fitness and weight loss, and people assume that physical activity can lose weight as a fact.
Also KOLs (Key Opinion Leader) usually encourage people to drink more water for higher metabolic levels.
Although I haven't seen any actual data for proving it, some papers have pointed the effect of drinking a lot of water on body weight [@vinu2013].
At the same time, the effect of physical activity to lose weight is also different from person to person.

@@

The data is obtained from a website [UCI Machine Learning Repository](https://archive.ics.uci.edu/dataset/544/estimation+of+obesity+levels+based+on+eating+habits+and+physical+condition) together with a paper as a source, which have collected the survey results of individuals from the countries of Mexico, Peru and Colombia on their eating habits and physical condition, as well as their estimation of obesity levels.
The original dataset contains 17 variables and 2111 observations, labeled with the class variable Obesity Level which has 7 categories [@PALECHOR2019104344].
The variables include gender, age, height, weight, family history, eating habits (high caloric food, vegetables, main meals, snacks), smoking, alcohol consumption, water consumption, physical activity, technological devices usage, commute transportation, and their obesity level according to the BMI of each observation but BMI value is not in the dataset directly.
Over weight level 1 refers to people whose BMI is in the range of 25-27.5, while weight level 2 is 27.5-30.
But these 2 levels couldn't be found on CDC website.
Obesity is frequently subdivided into categories: Class 1: BMI of 30 to \< 3; Class 2: BMI of 35 to \< 40; Class 3: BMI of 40 or higher.

@@

Several questions will be explored during the following analysis.

1.  If higher frequency of physical activity and drinking more water could keep people away from obesity.
2.  Whether there is a relationship between alcohol consumption frequency and obesity level.
3.  What variables/risk factors could be good predictors for obesity level, and what machine learning model could be fit and make good predictions for new given data.

@@

Then simple linear model will be applied to explore the relationship between variables and outcomes (both categorical and numeric), and proper predictors will be pick out from the variables.
And then machine learning models will be chosen and fitted to the dataset and each performance metric will be determined and compared for reporting the "best" model among the choices based on reality.

@@

The dataset presents the results of a survey on the topic of obesity including eating habits and personal physical conditions and answers to behavioral questions.
The paper have pointed out that 23% of data are collected by themselves directly from users through a web platform while 77% of the data is synthesized using the Weka tool and the SMOTE filter.

@@

There is no missing value in this dataset.

@@

First, for exploring the correlation between transportation type and physical activity frequency, transportation types were converted into numbers indicating the energy consuming levels from 1 (lowest automobile) to 4 (highest walking).
Then correlation function was used to calculate the coefficient for these two variable but it was pretty low (0.0059).

Then, linear model for BMI is fit by both water consumption and physical activity.
And the result indicates that there is a connection effect between them two, which means that both of them contribution to obesity control.

Also, single indicator, alcohol consumption is used and BMI as an outcome to fit linear model.
It shows that if an individual consumes alcohol sometimes than never, his BMI increase 3.98.
But the correlations are all negative for level 3 and 4, which means consuming more alcohol may decrease BMI.

Although age distribution is not so balanced in the dataset, the linear model of obesity by gender and age is still tired to explored.
So simple linear model is fit to the dataset using gender and age as independent variables.
The result indicates the interactions between them.
It is a factor out of my expectation.
And this interaction is considered to be included for next machine learning models.

In the last model selected is all the variables that I think may have effect on BMI as predictors and BMI as outcome.

Since it is quite obvious that family history is an important factor for obesity, and also some reasons how family obesity could influence the kids involved have been mentioned above in the introduction part, it is skipped for this basic exploratory analysis here, but will be included in the following part.

@@

Table 1.
Summary of each machine learning model applied to the data and its metric value comparison.

+----------------------------------+---------------------------------------------------------------+---------------+------------------------------+
| Model                            | Predictor                                                     | Metric Value  | Metric Value on testing data |
+==================================+===============================================================+===============+==============================+
| Linear null                      | \-                                                            | RMSE=8.01     | RMSE=7.99                    |
+----------------------------------+---------------------------------------------------------------+---------------+------------------------------+
| Linear regression 1              | Age, History, Alcohol, MTRANS                                 | RMSE=0.62     | RMSE=6.54                    |
+----------------------------------+---------------------------------------------------------------+---------------+------------------------------+
| Linear regression 2              | All                                                           | RMSE=6.29     | RMSE=3.81                    |
+----------------------------------+---------------------------------------------------------------+---------------+------------------------------+
| LASSO model                      | All                                                           | RMSE=6.34     | RMSE=6.47                    |
|                                  |                                                               |               |                              |
|                                  |                                                               | l=0.023       |                              |
+----------------------------------+---------------------------------------------------------------+---------------+------------------------------+
| Random forest                    | All                                                           | RMSE=3.35     | RMSE=3.97                    |
+----------------------------------+---------------------------------------------------------------+---------------+------------------------------+
| Logistic null                    | \-                                                            | Accuracy=0.17 | Accuracy=0.15                |
+----------------------------------+---------------------------------------------------------------+---------------+------------------------------+
| Discriminant analysis            | Gender, Age, History, Water, Alcohol, FAF, MTRANS, Gender_Age | Accuracy=0.50 | Accuracy=0.50                |
+----------------------------------+---------------------------------------------------------------+---------------+------------------------------+
| Multinominal logistic regression | All                                                           | Accuracy=0.55 | Accuracy=0.50                |
+----------------------------------+---------------------------------------------------------------+---------------+------------------------------+

The above table has listed all of the the models have been chosen to fit the data together with the null models.

When BMI is used as a continuous outcome, linear regression models can be applied to the data.
Compared to the linear null model, all of the ML models perform better according to RMSE value.
Since it is reasonable that physical activity is a potential factor that may influence the obesity level, although linear model with all variables and interactions has higher RMSE, regression model 2 that picks all of the predictors is better.
Then since random forest model selects all predictors and RMSE value is relatively lower among all the linear models, RF model is the choice for linear model.

@@

![](images/Observed-Predicted%20.png){width="2100"}

As for obesity level as a categorical outcome, logistic models are fit to the data.
And accuracy is used as the performance metric for this evaluation.
First, both the models have better performance than null model.
And according to the accuracy, multinominal logistic regression performs better.

Because random forest model is chosen between linear regression models and multinominal logistic regression between logistic models, test data is used to assess the quality of these two models separately.
And the new RMSE on test data for random forest model is 3.97, and it is 3.35 when the model is trained and cross validation is applied.
And the new accuracy is 0.50 when it is used on test data which is the same as it is trained.
The metric value difference is a sign that both of the models are overfitting.

So all of the models are assessed by making predictions test data in the table above and new metric values are collected again.
For linear regression 1, the RMSE on testing data is 6.54, which is much higher than that on training data.
So this model is not doing prediction well.
The new RMSE for linear regression 2 is 3.81 lower than 6.29 on training data, which is a sign of good prediction.
Besides, the new RMSE of LASSO model is 6.47 which is slightly higher than training.
Discriminant analysis has the same accuracy for both training and testing data.
Based on all the result of the machine learning I have used on dataset, I think linear regression model with LASSO regularization is a good choice among them.
Also, discriminant analysis has good performance on predicting in testing data.

@@

I chose a dataset that summarize an investigation result done in South American countries about individual physical behaviors and the classification of obesity level based on the information of height and weight obtained from the investigation.
I had some questions to explore and did data analysis on this dataset.
Samples were collected randomly, and some of the data was synthetic.
The result of data analysis indicate that both environmental factors and personal lifestyle habits can affect the degree of obesity according to bmi as a measurement.
It could provide some suggestions from this project that, although BMI might change with aging regard to gender, if someone can do more physical exercise, drink more water and consume less alcohol, it will reduce the risk for getting obesity.
And relevant departments and organizations can also provide corresponding recommendations and appropriate publicity to increase the awareness of obesity risk factors among residents of these countries and help them develop a healthy lifestyle.

@@

For machine learning model fitting, I found the best predictors for BMI index based on the value of some performance metrics after trying both linear regression and classification models, which were gender, age, family history, water drinking amount, frequency of physical activity, alcohol consumption frequency and commute transportation, and all of the categories were used in the analysis.
Although the machine learning models that I used for fitting the dataset are different types and their metrics and not be compared directly, the assessment of model quality using testing data could give a clue for the model selection.
Considering all the results of performance metrics, I finally chose liner regression model with LASSO regularization with all predictors plus BMI as the outcome.
The reason was that RMSE calculated of predicted MBI values on testing data split from cleaned dataset was as same as the result when it was trained.
Also, regularization was applied by LASSO model to reduce overfitting at the same time.

@@

BMI is a fast, low cost and easy-to-operate measurement for obesity rated in a population.
It applies to most of the people and could fit large population groups better than measuring body fat content.
I think my analysis has include two different types of model according to the type of outcome and each type include 2 models, which can provide more information and the result can be shown in different aspects as well as more choices were provided.
Also, interactions between predictors were found and included in the model fitting process.
Both predictors with and without interactions were all fit and compared with each other.
And LASSO model was used for decreasing overfitting.

First, there is a problem with my outcome, BMI, which [could be misleading when being used as an indicator for obesity](https://www.yalemedicine.org/news/why-you-shouldnt-rely-on-bmi-alone), because BMI value could be misleading because a person with lots of muscle and minimal body fat can have the same BMI as a person with obesity who has much less muscle.
The other reason is that BMI was invented based on European white men and didn’t take into account that a person’s body fat also tends to vary depending on their sex, race, and ethnicity.
So it will be better to take fat content into consideration when doing analysis for obesity.
But it is not included.
The difficulty for obtaining fat content might be that it consumes time to measure with specific machine.
I did tried to do tuning for random forest model but I failed for running it successfully, so this could be a limitation that I have.
And maybe there could be more metrics that I can also calculate but I only chose 1 for each model.
Furthermore, chatGPT told me that even if testing result has a sign for well predicting, there still could be other factors for further assessment such as F1-score and so on.
But base on my understanding, I don't know what other methods I could choose from so I just stopped here.
